---
title: "Traffic Stops: Predicting Searches & Testing for Bias"
subtitle: "A predictive and causal look at the Stanford Open Policing Project data"
format:
  html:
    default
  pdf:
    default
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(lubridate)
library(tidymodels)
library(bonsai)
library(xgboost)
library(marginaleffects)

set.seed(10)
```

# Question

Every good data science project starts with a clear question. We're
going to ask two, because this dataset supports two very different
*kinds* of question:

1. **Predictive question:** Using only information available at the
   moment a car is pulled over (time of day, location, reason for the
   stop, driver demographics), can we predict whether the officer will
   search the vehicle?

2. **Causal question:** Does reduced visibility at night *cause*
   officers to search drivers of different races at different rates?
   This is known in the research literature as the **"veil of
   darkness"** test — if officers can't clearly see a driver's race in
   the dark, but stop/search decisions still differ between day and
   night in a race-dependent way, that's evidence of bias in
   *deciding whether to stop someone in the first place*.

Notice these are different questions. The first asks "can we guess the
outcome?" The second asks "does one thing *cause* another?" A model
that predicts well is not automatically a model that explains cause
and effect.

# Wisdom

Before touching any code, it helps to know what we're working with and
why anyone cares.

The [Stanford Open Policing Project](https://openpolicing.stanford.edu/)
has collected and standardized over 200 million traffic stop records
from police departments across the United States, specifically so that
researchers can study patterns in how stops, searches, and citations
are carried out.

A few important background facts:

- Not every stop leads to a search. Officers decide, based on
  suspicion, whether to search a vehicle after stopping it.
- The "veil of darkness" idea (Grogger & Ridgeway, 2006) is a clever
  natural experiment: near sunset, the amount of daylight changes from
  day to day and week to week for reasons that have nothing to do with
  traffic or crime (it's just the Earth's orbit). So if you compare
  stops that happen at the *same clock time* but on days where that
  time happens to be "dusk" vs. "still daylight," you get something
  close to a random assignment of "could the officer see the driver's
  race clearly or not."
- This is sensitive subject matter. We are not trying to prove that
  any individual officer is biased — we're using statistics to look
  for *aggregate patterns* that are worth further investigation.

# EDA (Exploratory Data Analysis)

"EDA" means: before building any model, just look at the data. Get a
feel for it. Plot things. Notice anything weird.

```{r}
#| label: load-data
#| message: false

data_url <- "https://stacks.stanford.edu/file/druid:yg821jf8611/yg821jf8611_ri_statewide_2020_04_01.csv.zip"
local_zip <- "ri_statewide.csv.zip"

if (!file.exists(local_zip)) {
  download.file(data_url, destfile = local_zip, mode = "wb")
}

unzip(local_zip, exdir = ".", overwrite = TRUE)
csv_file <- list.files(pattern = "\\.csv$")[1]
stops <- read_csv(csv_file, show_col_types = FALSE)

glimpse(stops)
```

```{r}
#| label: clean-data

# Extract hour from time column using lubridate::hour()
# Keep only rows where we have all the info we need
# Convert categorical columns to factors
stops_clean <- stops |>
  filter(!is.na(search_conducted), !is.na(subject_race), !is.na(time)) |>
  mutate(
    hour = as.integer(hour(time)),  # Convert to integer to avoid type mismatch
    search_conducted = factor(search_conducted, levels = c(FALSE, TRUE),
                              labels = c("No", "Yes")),
    subject_race = factor(subject_race),
    subject_sex = factor(subject_sex),
    outcome = factor(outcome)
  ) |>
  select(hour, subject_race, subject_sex, outcome, search_conducted, date) |>
  drop_na()

cat("Cleaned data has", nrow(stops_clean), "rows\n")
cat("Search rate:", 
    round(mean(stops_clean$search_conducted == "Yes"), 3), "\n")
```

Let's look at how search rates vary by hour of day.

```{r}
#| label: search-rate-by-hour
#| echo: true

stops_clean |>
  group_by(hour) |>
  summarize(search_rate = mean(search_conducted == "Yes")) |>
  ggplot(aes(x = hour, y = search_rate)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = "Search Rate by Hour of Day",
    subtitle = "Rhode Island statewide police stops, 2005-2015",
    x = "Hour of Day (24-hour clock)",
    y = "Proportion of Stops Resulting in a Search"
  ) +
  theme_minimal()
```

And here's search rate broken down by driver race.

```{r}
#| label: search-rate-by-race
#| echo: true

stops_clean |>
  group_by(subject_race) |>
  summarize(search_rate = mean(search_conducted == "Yes"),
            n = n()) |>
  filter(n > 100) |>
  ggplot(aes(x = fct_reorder(subject_race, search_rate), y = search_rate)) +
  geom_col(fill = "darkorange") +
  coord_flip() +
  labs(
    title = "Search Rate by Driver Race",
    subtitle = "Raw rates differ across groups — but this ignores other factors",
    x = NULL,
    y = "Proportion of Stops Resulting in a Search"
  ) +
  theme_minimal()
```

That second plot shows exactly why we need *models*, not just raw
averages: race is correlated with lots of other things, so a raw
difference in rates doesn't tell us what causes the difference.

# Justice

"Justice" here means: set up the modeling fairly and correctly,
before we get excited about results.

```{r}
#| label: split-data

# Simple random split: 75% training, 25% test
set.seed(10)
train_idx <- sample(1:nrow(stops_clean), 
                    size = floor(0.75 * nrow(stops_clean)))

stops_train <- stops_clean[train_idx, ]
stops_test  <- stops_clean[-train_idx, ]

cat("Training set:", nrow(stops_train), "rows\n")
cat("Test set:", nrow(stops_test), "rows\n")
```

```{r}
#| label: recipe

# Recipe: predict search_conducted from hour, race, sex, outcome
# step_dummy() converts categories to 0/1 columns that xgboost understands
stops_recipe <- recipe(
  search_conducted ~ hour + subject_race + subject_sex + outcome,
  data = stops_train
) |>
  step_dummy(all_nominal_predictors())
```

# Courage

"Courage" is where we actually fit the model and look at how well it
does.

```{r}
#| label: fit-model
#| echo: true

# Define the XGBoost model spec
xgb_spec <- boost_tree(
  trees = 300,
  tree_depth = 5,
  learn_rate = 0.05
) |>
  set_engine("xgboost") |>
  set_mode("classification")

# Combine recipe and model into a workflow
stops_workflow <- workflow() |>
  add_recipe(stops_recipe) |>
  add_model(xgb_spec)

# Fit the model to training data
# This will take a minute or two
stops_fit <- fit(stops_workflow, data = stops_train)
```

```{r}
#| label: evaluate-model
#| echo: true

# Make predictions on test data
predictions <- augment(stops_fit, new_data = stops_test)

# Compute ROC AUC
roc_result <- predictions |>
  roc_auc(truth = search_conducted, .pred_Yes)

print(roc_result)
```

**Plain-English translation:** an AUC around 0.6–0.7 means the model
has found genuine patterns — time of day, driver demographics, and
outcome type do carry real information about whether a search happens.

```{r}
#| label: model-metrics
#| echo: true

# Confusion matrix: what did the model predict vs. what actually happened?
conf_matrix <- predictions |>
  conf_mat(truth = search_conducted, estimate = .pred_class)

autoplot(conf_matrix, type = "heatmap")

# Sensitivity: of the actual searches, how many did we catch?
cat("Sensitivity (true positive rate):\n")
print(predictions |>
  sensitivity(truth = search_conducted, estimate = .pred_class))

# Specificity: of the actual non-searches, how many did we get right?
cat("\nSpecificity (true negative rate):\n")
print(predictions |>
  specificity(truth = search_conducted, estimate = .pred_class))
```

# Temperance

"Temperance" means: don't overstate what the model tells you. Translate
model output into plain statements about effects, not just predictions.

```{r}
#| label: marginal-effects
#| echo: true

# Instead of using marginaleffects::avg_slopes(), let's manually compute
# the effect of hour by predicting at hour=10 vs hour=20 (holding else constant)

# Create two versions of the test data: one at hour 10, one at hour 20
test_hour_10 <- stops_test |>
  mutate(hour = 10L)

test_hour_20 <- stops_test |>
  mutate(hour = 20L)

# Get predicted search probabilities at each hour
pred_at_10 <- predict(stops_fit, new_data = test_hour_10, type = "prob")$.pred_Yes
pred_at_20 <- predict(stops_fit, new_data = test_hour_20, type = "prob")$.pred_Yes

# Average effect: how much does search probability change from hour 10 to 20?
effect_per_hour <- mean(pred_at_20 - pred_at_10) / (20 - 10)

cat("Average change in search probability per hour:\n")
cat(round(effect_per_hour, 4), "\n")
cat("Interpretation: Each additional hour is associated with a", 
    round(effect_per_hour * 100, 2), "percentage point change\n")
```

**Now the causal piece: the veil of darkness test.**

```{r}
#| label: veil-of-darkness
#| echo: true

# Restrict to a narrow dusk window (hours 18 and 19)
# Hour 19 is darker on average than hour 18
dusk_window <- stops_clean |>
  filter(hour %in% c(18, 19)) |>
  mutate(is_dark = hour == 19)

cat("Dusk window sample size:", nrow(dusk_window), "\n")
cat("Dark (hour 19):", sum(dusk_window$is_dark), "stops\n")
cat("Light (hour 18):", sum(!dusk_window$is_dark), "stops\n")

# Split dusk data by darkness
dark_data  <- dusk_window |> filter(is_dark)
light_data <- dusk_window |> filter(!is_dark)

# Recipe for T-learner (we're holding hour fixed at dusk, 
# so we don't include it)
t_recipe <- recipe(search_conducted ~ subject_race + subject_sex + outcome,
                    data = dusk_window) |>
  step_dummy(all_nominal_predictors())

# Fit XGBoost separately for dark and light conditions
fit_group_model <- function(data) {
  workflow() |>
    add_recipe(t_recipe) |>
    add_model(xgb_spec) |>
    fit(data = data)
}

model_dark  <- fit_group_model(dark_data)
model_light <- fit_group_model(light_data)

# Predict search probability for every driver in dusk window
# under BOTH lighting conditions, regardless of what they actually experienced
cate_estimates <- dusk_window |>
  mutate(
    pred_if_dark  = predict(model_dark,  new_data = dusk_window, 
                            type = "prob")$.pred_Yes,
    pred_if_light = predict(model_light, new_data = dusk_window, 
                            type = "prob")$.pred_Yes,
    effect_of_darkness = pred_if_dark - pred_if_light
  )

# Average causal effect of darkness, by race
causal_by_race <- cate_estimates |>
  group_by(subject_race) |>
  summarize(
    avg_effect_of_darkness = mean(effect_of_darkness),
    sd_effect = sd(effect_of_darkness),
    n = n(),
    .groups = "drop"
  ) |>
  filter(n > 30)

print(causal_by_race)

# Visualize
causal_by_race |>
  ggplot(aes(x = fct_reorder(subject_race, avg_effect_of_darkness), 
             y = avg_effect_of_darkness)) +
  geom_col(fill = "mediumpurple") +
  geom_errorbar(aes(ymin = avg_effect_of_darkness - sd_effect,
                    ymax = avg_effect_of_darkness + sd_effect),
                width = 0.2) +
  coord_flip() +
  labs(
    title = "Causal Effect of Darkness on Search Probability",
    subtitle = "By driver race (veil of darkness test)",
    x = NULL,
    y = "Predicted change in search probability\n(dark hour vs. light hour)"
  ) +
  theme_minimal()
```

**Plain-English translation:** `avg_effect_of_darkness` estimates how
much a driver's predicted search probability would change *just from
reduced visibility*, holding their violation type and other traits
fixed. If this differs meaningfully by race — if darkness matters more
for one group's search probability than another's — that's consistent
with bias in *who gets stopped in the first place*.

# Humility

No model, however cool, tells the whole story. Here's what to keep in
mind:

- **The predictive model only shows correlation, not cause.** A good
  AUC means the model found patterns in officer behavior — it says
  nothing about whether the underlying decisions were justified.
- **The veil of darkness design is clever, but not airtight.** Our
  simplified split (just hours 18 and 19) is a teaching simplification.
  A real analysis would compute exact local sunset times per date and
  location, and test for confounders.
- **This is real people's data.** Every row is a real traffic stop
  involving real people. Findings about aggregate disparities are not
  claims about any individual officer or driver.
- **Missing data isn't random.** Different police departments record
  different fields with different completeness, so comparisons can be
  distorted by reporting differences.
- **A model that predicts search well is not a policy recommendation.**
  It's a diagnostic tool for researchers and journalists — exactly
  what the Stanford team designed this dataset for.
